---
title: "Simulating PDMPs with event rates bounded by a polynomial"
output: github_document
---

In most situations the PDMP rate will not be exactly polynomial but can be bounded by a polynomial. Consider logistic regression where the gradient function is

```{r}
library(ccpdmp)
# Return the partial derivative for a logistic regression
get_grad <- function(x, index){
  expa <- exp(X%*%x)
  phi_1 <- expa/(1+expa) - y

  grad <- rep(0, length(index))
  for(i in 1:length(index)){
    grad[i] <- sum(phi_1*X[,index[i]])
  }
  return(grad)
}
```

Following the notation from the paper this can be simulated using the bound $f(t) \leq f(0) + f'(0)t+f''(0)t^2/2 + M/6$ where $f'''(t)\leq |M|$. The paper gives an explicit form for this here we show how this polynomial may be used in simulating the logistic regression example. In addition to supplying the gradient function the user must also specify a function that evaluates the polynomial upper-bounding rate.

```{r}
## Return the rates for an upper-bounding 3rd order polynomial
return_rates <- function(x, theta, tau_grid, dnlogpi, rate_updates){

  tau_length <- length(tau_grid);
  n_rates <- length(rate_updates)

  rates_eval <- matrix(0, n_rates, tau_length)
  a <- X%*%x
  expa <- exp(a)
  da_dt <- X%*%theta

  # First 3 derivatives of phi(a) = log(1+exp(a)) - ya
  phi_1 <- expa/(1+expa) - y
  phi_2 <- expa/(1+expa)^2
  phi_3 <- -expa*(expa -1)/(expa+1)^3
  phi_4_bound <- 1/8

  for(i in 1:n_rates){
    ## Calculate f(t) and derivatives of f(t) for Taylor expansion
    partial_i <- rate_updates[i]
    f <- theta[partial_i]*sum(phi_1*X[,partial_i])
    f_1 <- theta[partial_i]*sum(phi_2*X[,partial_i]*da_dt)
    f_2 <- theta[partial_i]*sum(phi_3*X[,partial_i]*da_dt^2)
    f_3 <- sum(phi_4_bound*abs(X[,partial_i]*da_dt^3))

    # define the upper-bounding polynomial
    # p(t) = f(0) + t*f'(0) + t^2f''(0)/2 + M/6
    poly <- c(f_3/factorial(3), f_2/factorial(2), f_1, f)

    # Evaluate the upperbound at the tau_grid points
    rates_eval[i,] <- pracma::polyval(poly, tau_grid)
  }
  return(rates_eval)
}
```

Example run:

```{r}
generate.logistic.data <- function(beta, n.obs, siginv) {
  p <- length(beta)
  dataX <- mvtnorm::rmvnorm(n=n.obs*p,sigma = solve(siginv))
  vals <- dataX%*%beta
  generateY <- function(p) { rbinom(1, 1, p)}
  dataY <- sapply(1/(1 + exp(-vals)), generateY)
  return(list(dataX = dataX, dataY = dataY))
}
set.seed(1)
n_ev <- 1e3; p <- 3; n <- 100
beta <- c(c(-1.25, 0.5), rep(-0.4,p-2))
siginv <- diag(1, p,p)
siginv[1,2] <- siginv[2,1] <- 0.8
data <- generate.logistic.data(beta, n, siginv)
X <- data$dataX
y <- data$dataY

z <- zigzag(5e3, dnlogpi = get_grad, return_rates = return_rates,x0 = beta, poly_order = 3)
plot_pdmp(z, nsamples = 5e3, inds = 1:5e3)
```



## Local methods and efficient C++ implementations

More efficient versions of the samplers are also available though they require specifying the concave-convex decomposition using a C++ implementation. The code below defines a cpp function which takes in the current position and velocity (x, theta) data y and any additional data from a list and returns a matrix with concave convex decomposition. In this case the distribution is a Gaussian.

```{r}
library(RcppXPtrUtils)
local_rate <- cppXPtr("arma::vec get_rate(double t, arma::vec& t_old, arma::vec& x, arma::vec& theta,
                      const arma::vec& y, const List& Data, arma::uvec rate_inds, bool grad) {

                     // Standard header //
                     int num_terms = rate_inds.size(), rsize = 5;
                     double x_t, grd;
                     if(grad){
                        rsize += num_terms;
                     }
                     arma::vec res = arma::zeros(rsize);

                     // Evaluate the rate: (f_u, f_n, f_n', p, f) for the factor
                     // Return additional gradient terms if grad = TRUE

                     for(int i = 0; i < num_terms; i++){

                          int partial = rate_inds(i);

                          x_t = x(partial) + theta(partial)*(t-t_old(partial));
                          grd = x_t;
                          res(0) += theta(partial)*grd;

                          if(grad){
                          res(5 + i) = grd;
                          }
                     }
                     res(4) = res(0) + res(1) + res(3);
                     return(res);
                     }", depends = c("RcppArmadillo"))
```

Once this has been set up the practitioner may specify a choice of Factors which partition the parameter set and local updates indicating which factors need re-simulation once and event for the corresponding factor is triggered.

```{r}
set.seed(0)
d <- 1000
x0<- rnorm(d)
theta0 <- rnorm(d)
nmax <- 10^5
ref_rate <- 0.05
Datann <- list()

## Standard BPS -- A single factor consisting of all variables
Factors <- list(c(0:(d-1)))
Neighbourhoods <- list(c(0)) # just resimulate the Factor 0 (the only factor)

system.time({set.seed(1);bps_global <- bps_cpp(maxTime = 2, # Run time
                                               trac_coords = c(0,1,d-1), # Coords to track
                                               rate_f = local_rate, # Cpp function
                                               factors = Factors, # List of factors
                                               local_updates = Neighbourhoods, # List of factors to update
                                               Data = Datann, # Possible extra parameters
                                               y = y,         # Possible extra data
                                               nmax = d*3*10^2, # maximum number of events
                                               x0 = x0, # initial location
                                               theta0 = theta0, # initial velocity
                                               ref_rate = 1)})
length(bps_global$times)
plot_pdmp(bps_global, pch = '.',coords = c(1,2,3), nsamples = 1e3, inds = 1:1e4)

## local BPS -- 10 factors consisting of d/10 variables each
Factors <- lapply(0:(d/10-1), function(i) seq(from = 10*i, to = 10*i+9))
Neighbourhoods <- lapply(0:(d/10-1), function(i) c(i))

system.time({set.seed(1);bps_local <- bps_cpp(maxTime = 2, # Run time
                                               trac_coords = c(0,1,d-1), # Coords to track
                                               rate_f = local_rate, # Cpp function
                                               factors = Factors, # List of factors
                                               local_updates = Neighbourhoods, # List of factors to update
                                               Data = Datann, # Possible extra parameters
                                               y = y,         # Possible extra data
                                               nmax = d*3*10^2, # maximum number of events
                                               x0 = x0, # initial location
                                               theta0 = theta0, # initial velocity
                                               ref_rate = 1)})
length(bps_local$times)
plot_pdmp(bps_local, pch = '.',coords = c(1,2,3), nsamples = 1e3, inds = 1:1e4)
```

